Immutable Source-Mapped Abstract Syntax Tree:\ \, \\ A Design Pattern for Refactoring Engine APIs

### 2017-09-27 paper.tex
Removed flow-disrupting parentheses

Developers familiar with the HTML or XML DOM (Document Object Model) will find
this structure familiar.  The DOM is effectively an abstract syntax tree for
HTML/XML documents.

Deeper discussions on ASTs can be found in many introductory textbooks on
compiler construction, such as \cite{aho06compilers} and
\cite{torczon11engineering}.  Such books also discuss how parsers are
constructed, including how they can be modified to construct ASTs.

@@

To illustrate the first requirement, consider an example from Fortran.  The
Fortran language contains a number of input/output statements.  Two of these
are \textsc{print} and \textsc{write}: the statements \ttt{print *, "X"} and
\ttt{write (*,*) "X"} can be used interchangeably to print the string \ttt{X}
to standard output.  In general, the statement \ttt{print} \textit{fmt}\ttt{,}
\textit{expr} is equivalent to the statement \ttt{write (*,
}\textit{fmt}\ttt{)} \textit{expr}.  Most compilers type check and translate
them identically.  So, a compiler may choose to represent \textsc{print}
statements as \textsc{write} statements in its AST.  The GNU Fortran compiler
does this, for example.  However, the ability to convert \textsc{print}
statements to \textsc{write} statements actually makes for a useful
refactoring.  (Consider a program that uses \textsc{print} statements to write
to standard output, but the programmer later needs to use \textsc{write}
statements to write to a file instead.) To implement this refactoring, a
refactoring tool needs to be able to distinguish \textsc{print} statements from
\textsc{write} statements---a difficult task if they are represented
identically in the AST.

@@

Both Clang and the original refactoring engine in Xcode~3.0 modify source code
using offset-length information directly.  Notably, Clang offers a
\textit{RewriteRope} class---essentially, a custom string class that allows for
efficient mid-string insertions and
deletions\footnote{RewriteRope Class Reference: \url{http://clang.llvm.org/doxygen/classclang_1_1RewriteRope.html}}.
(It uses a B-tree structure inspired by, but not identical to, that described
in the original paper on Ropes~\cite{boehm95ropes}.)

>>>>

@@

To illustrate the first requirement, consider an example from Fortran.  The
Fortran language contains a number of input/output statements.  Two of these
are \textsc{print} and \textsc{write}: the statements \ttt{print *, "X"} and
\ttt{write (*,*) "X"} can be used interchangeably to print the string \ttt{X}
to standard output.  In general, the statement \ttt{print} \textit{fmt}\ttt{,}
\textit{expr} is equivalent to the statement \ttt{write (*,
}\textit{fmt}\ttt{)} \textit{expr}.  Most compilers type check and translate
them identically.  So, a compiler may choose to represent \textsc{print}
statements as \textsc{write} statements in its AST.  The GNU Fortran compiler
does this, for example.  However, onsider a program that uses \textsc{print} 
statements to write to standard output, but the programmer later needs to 
use \textsc{write}
statements to write to a file instead. Then the ability to convert \textsc{print}
statements to \textsc{write} statements actually makes for a useful
refactoring. To implement this refactoring, a
refactoring tool needs to be able to distinguish \textsc{print} statements from
\textsc{write} statements---a difficult task if they are represented
identically in the AST.

@@

Both Clang and the original refactoring engine in Xcode~3.0 modify source code
using offset-length information directly.  Notably, Clang offers a
\textit{RewriteRope} class---essentially, a custom string class that allows for
efficient mid-string insertions and
deletions\footnote{RewriteRope Class Reference: \url{http://clang.llvm.org/doxygen/classclang_1_1RewriteRope.html}}.
It uses a B-tree structure inspired by, but not identical to, that described
in the original paper on Ropes~\cite{boehm95ropes}.
